---
title: "AmNat analysis"
author: Vidar Grøtan (vidar.grotan@ntnu.no)
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
```


```{r}
library(popbio)
library(jtools)
library(lsei)
library(TMB)
library(glmmTMB)
```

### Utility functions  

Some functions for calculating matrices based on vital rates etc

```{r, echo = TRUE, include = TRUE}
# calculate annual Lefkovitch-matrices


#' Calculate annual Lefkovitch-matrices
#'
#' Assumes recruitment and survival rates of one or both sexes
#'
#' @param rec recruitment rate
#' @param surv survival rate
#' @param K number of age classes
#' @param year used for naming years in the returned array

#'
#' @return An array of Lefkovitch-matrices, K x K x number of years
#' @export
#'
#' @examples
construct.Lij_t <- function(rec, surv, K, year = NULL){
  Lij_t <- array(0, c(K,K,nrow(rec)))
  if (is.null(year)){
    year <- rownames(rec)
  }
  dimnames(Lij_t) <- list(1:K, 1:K, year)

  Lij_t[1, , 1:nrow(rec)] <- t(rec)
  for (i in 2:K){
    Lij_t[i, i-1, ] <- surv[,i-1]
  }
  Lij_t[K, K, ] <- surv[,K]

  Lij_t
}

#' Calculate lambda given a matrix or array of matrices (e.g. a matrix per year)
#'
#' Assumes recruitment and survival rates of one or both sexes
#'
#' @param Lij_t An array of transition matrix or a single matrix
#'
#' @return a vector of lambdas (if Lij_t is an array)
#' @export
#'
#' @examples
calcLambda <- function(Lij_t){
  if (class(Lij_t) == "array"){
    lambda <- apply(Lij_t, 3, function(x) popbio::eigen.analysis(x)$lambda1)
  }else{
    lambda <- popbio::eigen.analysis(Lij_t)$lambda1
  }
  lambda
}

# calculate lambda by adding a constant factor across years to recruitment
# to adjust for immigration / emigration
# mean log(lambda) across years should be = 0
fImmAdj <- function(frac){
  Lij_t <- construct.Lij_t(rec*frac, surv, K)
  lambda <- calcLambda(Lij_t)
  (mean(log(lambda)))^2
}
```


### Data

Individual specific data on survival and recruitment obtained from:
https://pure.knaw.nl/portal/en/publications/evidence-for-r-and-k-selection-in-a-wild-bird-population-a-recipr/datasets/

Estimates of number of individuals in different age classes were obtained as part of the statistical analysis of the following paper: 
Marlène Gamelon, Vidar Grøtan, Steinar Engen, Eirin Bjørkvoll, Marcel E Visser, Bernt‐Erik Sæther. 2016. Density dependence in an age‐structured population of great tits: identifying the critical 3 age classes. Ecology 97: 2479-2490. https://doi.org/10.1002/ecy.1442

```{r}
r.K.GT <- read.delim("r.K.dat.GT.HV.sel.txt",header=T)
N_ind <- with(r.K.GT, tapply(FemSurvYN, list(Year, Age), length))
R <- with(r.K.GT, tapply(Tot.recr.loc/2, list(Year, Age), sum))
varR <- with(r.K.GT, tapply(Tot.recr.loc/2, list(Year, Age), var))/N_ind
S <- with(r.K.GT, tapply(FemSurvYN, list(Year, Age), sum))

# posterior mean number of individuals in age classes 1 to 4 - obtained from 
# statistical analysis in Gamelon et al. 2016: https://doi.org/10.1002/ecy.1442
N_estimate <- structure(list(V1 = c(68.971, 102.448, 45.367, 115.439, 60.7195, 
43.1125, 99.1465, 73.4635, 56.452, 68.173, 89.4935, 23.951, 62.624, 
79.831, 103.212, 21.2455, 58.4965, 106.2875, 49.4575, 77.584, 
59.658, 52.151, 95.655, 53.742, 61.838, 62.545, 35.989, 94.168, 
41.848, 104.578, 49.5685, 103.1345, 38.787, 83.1135, 72.472, 
58.154, 70.5755, 78.774, 110.6425), V2 = c(56.1065, 36.5505, 
52.8675, 29.7705, 29.444, 23.785, 26.9015, 44.0145, 39.632, 28.4215, 
34.7725, 21.592, 11.18, 31.083, 51.4745, 43.5145, 9.716, 36.671, 
52.2175, 24.556, 36.097, 22.53, 31.0215, 28.7715, 26.171, 38.4325, 
26.352, 22.763, 34.192, 28.6385, 23.8745, 28.905, 29.266, 24.57, 
47.98, 24.834, 29.628, 29.408, 48.5715), V3 = c(26.3685, 31.1425, 
22.027, 35.7435, 11.0045, 14.1065, 16.7035, 13.9705, 26.1595, 
22.7305, 13.8395, 10.034, 10.717, 6.163, 19.5665, 24.4205, 23.366, 
6.686, 18.794, 23.2355, 12.497, 13.703, 14.3405, 11.1175, 15.6245, 
17.432, 16.6485, 18.7265, 10.2445, 23.0065, 8.23, 16.261, 12.608, 
21.4655, 21.574, 19.381, 15.6875, 14.9885, 20.5195), V4 = c(21.6315, 
18.312, 22.267, 20.6565, 13.7775, 7.688, 10.76, 10.2, 10.4535, 
16.044, 13.0495, 5.0955, 5.4275, 6.144, 5.9075, 10.64, 15.483, 
17.225, 8.3415, 9.4785, 11.1865, 7.9625, 11.429, 7.158, 10.463, 
13.4805, 10.006, 14.499, 11.1755, 11.234, 8.0215, 10.235, 10.2, 
13.452, 20.5325, 14.3705, 13.5135, 10.1985, 13.4765)), class = "data.frame", row.names = c(NA, 
-39L))
# sum the number of individuals in age classes 3 and 4+ into age class 3+
N_estimate[,3] <- N_estimate[,3]+ N_estimate[,4]
N_estimate <- N_estimate[,-4]
# remove 2012 (individual data only until 2011)
N_estimate <- N_estimate[-nrow(N_estimate),]

indata <- list(TT = as.integer(nrow(R)), K = 3,
               N_ind = data.matrix(N_ind), R = data.matrix(R),
               S = data.matrix(S), N_estimate = data.matrix(N_estimate), varR=data.matrix(varR))

```


The overall structure of the data used as the starting point for the statistical analysis in the paper. 
```{r, echo = TRUE, include = TRUE}

# TT : number of years with data
# K : number of adult age classes/stages (in this case age 1, 2, >=3 - hereafter 1, 2, 3)
# N_ind : number of data points (= number of females per age per year) used for calculating:
# R : mean number of recruits (entry in year t: offspring produced in year t that survives until year t+1) per age class per year divided by 2 (assuming 1:1 sex ratio) : female recruits
# S : mean adult survival per year per age class (entry in year t: survival from year t to t+1)
# N_estimate : estimate of N in different age classes from Gameleon et al
# VarR: Variance of number of female recruits (per female ) calculated for R given definition above

# load the data given in the script, show the structure 
str(indata)
```


```{r}
K <- indata$K
rec <- (indata$R/indata$N_ind) 
surv <- indata$S/indata$N_ind
K <- indata$K
Ntot <- indata$N_estimate
```


### Calculate $\lambda$ from age-specific vital rates and adjust for immigration

```{r}
# construct Lij_t based on rec and surv data
Lij_t_raw <- construct.Lij_t(rec, surv, K)

# calculate lambda based on Lij_t - unadjusted for immigration
lambda_raw <- calcLambda(Lij_t_raw)
loglambda_raw <- log(lambda_raw)

immfactor <- optimize(fImmAdj, lower=0, upper=10)$minimum
immfactor
immfactor/(immfactor+1) # immigrants relative to local recruits

```


```{r}
# yearly transition matrices accounting for immigration
Lij_t <- construct.Lij_t(rec*immfactor, surv, K)
# calculate lambda based on Lij_t - accounting for immigration
lambda <- calcLambda(Lij_t)
loglambda <- log(lambda)
```



#### calculate weights

```{r}
# loglambda vs total pop size
N123 <- rowSums(Ntot)
f <- lm(loglambda ~ N123)
summ(f, confint = TRUE, digits = 8)
effect_plot(f, pred = N123, interval = TRUE, plot.points = TRUE)

# regression of loglambda ~ N1 + N2 + N3
N1 <- Ntot[,1]
N2 <- Ntot[,2]
N3 <- Ntot[,3]

fm <- lm(loglambda ~ N1 + N2 + N3)
summ(fm, confint = TRUE, digits = 8)
#plot_summs(fm, scale = TRUE, plot.distributions = TRUE, inner_ci_level = .95)
effect_plot(fm, pred = N1, interval = TRUE, plot.points = TRUE)
effect_plot(fm, pred = N2, interval = TRUE, plot.points = TRUE)
effect_plot(fm, pred = N3, interval = TRUE, plot.points = TRUE)
#

# gamma and weights b_i
gammab_i <- coef(fm)[-1]
b_i <- gammab_i/gammab_i[1]
gamma <- - gammab_i[1] # b_1 = 1 !
gamma
b_i
```

We set positive slopes to 0:

```{r}
# drop N3 from model fit
fm2 <- lm(loglambda ~ N1 + N2)
summ(fm2, confint = TRUE, digits = 8)
#plot_summs(fm, scale = TRUE, plot.distributions = TRUE, inner_ci_level = .95)
effect_plot(fm2, pred = N1, interval = TRUE, plot.points = TRUE)
effect_plot(fm2, pred = N2, interval = TRUE, plot.points = TRUE)
#

# gamma and weights b_i
gammab_i <- c(coef(fm2)[-1], 0)
b_i <- gammab_i/gammab_i[1]
gamma <- - gammab_i[1] # b_1 = 1 !
gamma
b_i
```


The weights should sum to K (the number of age classes):
```{r}
# keep a copy of the weights
b_i_raw <- b_i
# weights should sum to 3
b_i <- b_i * K / sum(b_i)
b_i_raw
b_i
sum(b_i)

```

### Regression of log(lambda) on weighted total N (step 1) 

```{r}
Nw <- as.vector(b_i %*% t(Ntot))
plot(loglambda~Nw)
lm.Nw <- lm(loglambda ~ Nw)
effect_plot(lm.Nw, pred = Nw, interval = TRUE, plot.points = TRUE)
summ(lm.Nw, confint = TRUE, digits = 8)
```

Statistics on Nw:
```{r}
mean(Nw)
var(Nw)
sd(Nw)/mean(Nw)

```
```{r}

#png("Plot2.png", width = 16, height = 10, units = 'cm', res = 300)
#pdf("Plot2.pdf", width = 16/2.54, height = 10/2.54)
par.old <- par(no.readonly = TRUE)
par(las = 1, 
    mar=c(0.8, 1.5, 0.1, 0.1), 
    omi=c(0.8, 0.8, 0.03, 0.03))

lm.Nw <- lm(loglambda ~ Nw)
newx <- seq(min(Nw), max(Nw), length.out=100)
preds <- predict(lm.Nw, newdata = data.frame(Nw=newx), interval = 'confidence')

plot(loglambda ~ Nw, type="n", axes = FALSE, ann = FALSE, xlim=c(70, 250), ylim=c(-1.6, 0.95))
points(loglambda ~ Nw, pch=20)
lines(newx, preds[ ,1], lwd = 2, col = 'black')
#add dashed lines for confidence bands
#lines(newx, preds[ ,3], lty = 2, col = 'darkgrey', lwd=2)
#lines(newx, preds[ ,2], lty = 2, col = 'darkgrey', lwd=2)
# polygon

polygon(c(rev(newx), newx), c(rev(preds[ ,3]), preds[ ,2]), col = rgb(0.7,0.7,0.7,0.4) , border = NA)
axis(1, at = seq(50, 250, by=25))
axis(2, at = seq(-2, 1, by=0.5))
#dev.off()
par(par.old) 

```


### Graph of weighted N and N versus time

```{r}

# WEIGHTED N AND UNWEIGHTED N VERSUS TIME
#png("Plot1.png", width = 16, height = 10, units = 'cm', res = 300)
#pdf("Plot1.pdf", width = 16/2.54, height = 10/2.54)
par.old <- par(no.readonly = TRUE)
par(las = 1, 
    mar=c(0.8, 1.5, 0.1, 0.1), 
    omi=c(0.8, 0.8, 0.03, 0.03))

par(las=1)
plot(1974:2011, Nw, type="n", axes=FALSE, ann = FALSE, ylim=c(50, 250), xlim=c(1974, 2011))
#title(xlab="Year", ylab="Population size")
lines(1974:2011, Nw)
lines(1974:2011, rowSums(Ntot), lty=2)
axis(1, at= seq(1970, 2015, by=5))
axis(2, at= seq(0, 350, by=50))
#dev.off()
par(par.old) 

```

### Step 2 separate linear regressions


```{r}
#################################################################################
#################################################################################
# step 2 : separate linear regressions
#
# input :   b_i : weights
#           gamma : densdep
#################################################################################
#################################################################################


lmdat <- data.frame(rec,surv,Nw)
colnames(lmdat)[1:(K*2)] <- paste0(rep(c("r", "s"), each = K), 1:K)

# recruitment * immfactor!
fr1 <- lm(r1*immfactor ~ Nw, data = lmdat)
summ(fr1, confint = TRUE, digits = 8)
pfr1 <- effect_plot(fr1, pred = Nw, interval = TRUE, plot.points = TRUE, data = lmdat)

fr2 <- lm(r2*immfactor ~ Nw, data = lmdat)
summ(fr2, confint = TRUE, digits = 8)
pfr2 <- effect_plot(fr2, pred = Nw, interval = TRUE, plot.points = TRUE, data = lmdat)

fr3 <- lm(r3*immfactor ~ Nw, data = lmdat)
summ(fr3, confint = TRUE, digits = 8)
pfr3 <- effect_plot(fr3, pred = Nw, interval = TRUE, plot.points = TRUE, data = lmdat)

# survival
fs1 <- lm(s1 ~ Nw, data = lmdat)
summ(fs1, confint = TRUE, digits = 8)
pfs1 <- effect_plot(fs1, pred = Nw, interval = TRUE, plot.points = TRUE, data = lmdat)

fs2 <- lm(s2 ~ Nw, data = lmdat)
summ(fs2, confint = TRUE, digits = 8)
pfs2 <- effect_plot(fs2, pred = Nw, interval = TRUE, plot.points = TRUE, data = lmdat)

fs3 <- lm(s3 ~ Nw, data = lmdat)
summ(fs3, confint = TRUE, digits = 8)
pfs3 <- effect_plot(fs3, pred = Nw, interval = TRUE, plot.points = TRUE, data = lmdat)
```

```{r}

#setEPS()
#postscript("fig3.eps", horizontal = FALSE, paper = "a4")

#png("Plot3.png", width = 16, height = 10, units = 'cm', res = 300)

#pdf("Plot3.pdf", width = 16/2.54, height = 10/2.54)
par.old <- par(no.readonly = TRUE)

par(mfrow=c(2,3), las = 1, mar=c(0.8, 0.8, 0.1, 0.1), omi=c(0.5, 0.8, 0.1, 0.1))

plotFunc <- function(f, YLIM, yAT, XLIM = c(65, 250), xlabels = TRUE, ylabels = TRUE){
  preds <- predict(f, newdata = data.frame(Nw=newx), interval = 'confidence')
  plot(f$model[,1] ~ f$model[,2], type = "n", axes = FALSE, ann = FALSE, xlim=XLIM, ylim=YLIM, pty="s")
  points(f$model[,1] ~ f$model[,2], pch=20)
  lines(newx, preds[ ,1], lwd = 2, col = 'black')
  polygon(c(rev(newx), newx), c(rev(preds[ ,3]), preds[ ,2]), col = rgb(0.7,0.7,0.7,0.4) , border = NA)
  if (xlabels) axis(1, at = seq(50, 250, by=50)) else axis(1, labels=FALSE, at = seq(50, 250, by=50))
  if (ylabels) axis(2, at = yAT) else axis(1, at = yAT, labels = FALSE)
}
plotFunc(fr1, YLIM = c(0, 2.5), yAT = seq(-0.5, 2.5, by=0.5), xlabels = FALSE)
plotFunc(fr2, YLIM = c(0, 2.5), yAT = seq(-0.5, 2.5, by=0.5), xlabels = FALSE, ylabels = FALSE)
plotFunc(fr3, YLIM = c(0, 2.5), yAT = seq(-0.5, 2.5, by=0.5), xlabels = FALSE, ylabels = FALSE)

YLIM_s <- c(0, 1)
plotFunc(fs1, YLIM = YLIM_s, yAT = seq(-0.5, 2.5, by=0.25))
plotFunc(fs2, YLIM = YLIM_s, yAT = seq(-0.5, 2.5, by=0.25), ylabels = FALSE)
plotFunc(fs3, YLIM = YLIM_s, yAT = seq(-0.5, 2.5, by=0.25), ylabels = FALSE)

#dev.off()
par(par.old) 

#dev.off()

```



### Step 3 : estimate the environmental covariance matrix

Previous steps : models fitted to recruitment and survival, step 3 estimates the environmental covariance matrix.

First step is to calculate a covariance matrix from residuals - will include both environmental and demographic contributions:

```{r}
##### step 3 : estimate / adjust the environmental covariance matrix
##########################################################################

eps <- cbind(resid(fr1), resid(fr2), resid(fr3), resid(fs1), resid(fs2), resid(fs3))
#eps <- cbind(resid(fr1)/immfactor, resid(fr2)/immfactor, resid(fr3)/immfactor, resid(fs1), resid(fs2), resid(fs3))

C_ij_dem <- var(eps)
print(C_ij_dem, digits = 8)
```

When we calculate the demographic variance and subtract from the diagonal (eq. 4d) we get a matrix that is not positive definite. 

First calculate the demographic variance for vital rates (average over years) - D is ordered as recruitment[age:1-3] and survival[age:1-3]

We assume no demographic individual covariance between individual survival and reproduction within age classes.
```{r, error=TRUE}

##### step 3 : estimate / adjust the environmental covariance matrix
##########################################################################
N <- indata$N_ind
sd1 <- mean(predict(fs1)*(1-predict(fs1))/N[,1])
sd2 <- mean(predict(fs2)*(1-predict(fs2))/N[,2])
sd3 <- mean(predict(fs3)*(1-predict(fs3))/N[,3])
#varR <- (with(r.K.GT, tapply(RecruitsLocal/2, list(Year, Age), var))/N)
varR <- indata$varR#/indata$N_ind
varR <- varR * immfactor^2 # Immigration correction for recruits (dem var)
rd <- colMeans(varR)

D <- c(rd, sd1, sd2, sd3)
D

Dmat <- cbind(varR, predict(fs1)*(1-predict(fs1))/N[,1], predict(fs2)*(1-predict(fs2))/N[,2], predict(fs3)*(1-predict(fs3))/N[,3])
```


```{r}
C_ij <- C_ij_dem
diag(C_ij) <- diag(C_ij_dem) - c(rd, sd1, sd2, sd3)
# covariance matrix
print(C_ij, digits = 8)

```
This matrix is not positive definite
```{r, error=TRUE}
#chol(C_ij)
```


```{r}
#### summmary:###
# we have:
# - eps : matrix of residuals from fitted models, residuals because of env and dem stoch
# - eps : years in rows, r[1, 2, .., K], s[1, 2, .., K] : -> 2*K columns
# - We have the "demographic variance" for all observations based 
# on N (number of individual data points each year):

#### Aim: ####
# - obtain positive definite variance covariance matrix for eps
# - eps[year, 1:2*K] ~ MVN(0, SIGMA)
# SIGMA_t : D_t'ED_t,
# E is a positive definite variance unstructured covariance matrix for env stoch
# D_t is a diagonal matrix with dem var as calculated above on the 
# diagonal - varies from year to year

#### Method: ####
# - eps and D are data (dim = [1:T, 1:(2*K)])
# - estimate cholesky "factors" of E (gives positive definite E)
# - obtain E from cholesky factors
# for (t in 1:T)
# - D_t = diag(D[t,1:(2*K)]) # diagonal matrix
#- SIGMA_t = D_t' E D_t
# - eps[t, 1:(2*K)] ~ MVN(0, SIGMA_t)

covarEnvDem <- '
//
#include <TMB.hpp>

template<class Type>
Type objective_function<Type>::operator() () {
// data:
DATA_MATRIX(eps);
DATA_VECTOR(D);

PARAMETER_VECTOR(rd); // Dummy parameterization of correlation matrix
PARAMETER_VECTOR(logsds); // standard deviations for covariance matrix

vector<Type> sds = exp(logsds);
ADREPORT(sds);

using namespace density;
int i, j, k;
int NT = eps.cols();
int NY = eps.rows();

// Cholesky factor of Sigma
matrix<Type> L(NT,NT);
L.setIdentity();
k=0;

for(i=1;i<NT;i++)
{
  Type Norm2=L(i,i);
  for(j=0;j<=i-1;j++)
  {
    L(i,j) = rd(k++);
    Norm2 += L(i,j)*L(i,j);
  }
  for(j=0;j<=i;j++)
    L(i,j) /= sqrt(Norm2);
}

matrix<Type> RhoSigma = L * L.transpose();
ADREPORT(RhoSigma);

// Scale up correlation matrix
matrix<Type> E(NT,NT);
matrix<Type> Sigma(NT,NT);
for(int i2=0;i2<NT;i2++){
  for(j=0;j<NT;j++){
    E(i2,j) = RhoSigma(i2,j)*sds(i2)*sds(j);
    Sigma(i2,j) = E(i2,j);
  }
}
ADREPORT(E);

for(j=0;j<NT;j++){
 Sigma(j,j) = Sigma(j,j) + D(j);
}
ADREPORT(Sigma);

Type res=0;
vector<Type> u(NT);

for(j=0;j<NY;j++){
  u = eps.row(j);
  res += MVNORM(Sigma)(u);       // Evaluate negative log likelihod of env noise
}

return res;
}

'
write(covarEnvDem, file = "covarEnvDem.cpp")

```

```{r}
epsdata <- list(eps=eps, D = c(rd, sd1, sd2, sd3))
parameters <- list(rd=rep(0.1, choose(6,2)), logsds = log(sqrt(rep(0.02,6))))
TMB::compile("covarEnvDem.cpp")
dyn.load("covarEnvDem")
tmb_obj <- MakeADFun(epsdata, parameters, silent=T, DLL="covarEnvDem")
tmb_opt <- nlminb(tmb_obj$par,tmb_obj$fn,tmb_obj$gr, control=list(eval.max=5000, iter.max=5000))

```


```{r}
rep2 <- sdreport(tmb_obj, ignore.parm.uncertainty = TRUE)
est2 <- summary(rep2, "report")
C_ij_alt2 <- est2[rownames(est2) == "E",1] 
C_ij_dem_alt2 <- est2[rownames(est2) == "Sigma",1] 

```

#### ENVIRONMENTAL COVARIANCE MATRIX 
```{r}
C_ij_alt <- matrix(C_ij_alt2, ncol=K*2)
round(C_ij_alt, 8)
is.matrix(chol(C_ij_alt))
# eigen values
eigen(C_ij_alt)
# correlation matrix 
cov2cor(C_ij_alt)
```

#### TOTAL COVARIANCE MATRIX (ENV + DEM)

```{r}
C_ij_dem_alt <- matrix(C_ij_dem_alt2, ncol=K*2)
C_ij_dem_alt
is.matrix(chol(C_ij_dem_alt))
# correlation matrix 
cov2cor(C_ij_dem_alt)
```


### Input to Mathematica script 

```{r}
# A
A <- construct.Lij_t(rec=matrix(c(coef(fr1)[1], coef(fr2)[1], coef(fr3)[1]), nrow=1), surv=matrix(c(coef(fs1)[1], coef(fs2)[1], coef(fs3)[1]), nrow=1), K=3)[,,1]
Gamma <- -construct.Lij_t(rec=matrix(c(coef(fr1)[2], coef(fr2)[2], coef(fr3)[2]), nrow=1), surv=matrix(c(coef(fs1)[2], coef(fs2)[2], coef(fs3)[2]), nrow=1), K=3)[,,1]
Env <- C_ij_alt
```


```{r, echo=FALSE}
cat("A:", "\n")
print(A, digits=8)
cat("Gamma:", "\n")
print(Gamma, digits=8)
cat("Env:", "\n")
print(Env, digits=8)

```



```{r}
sessionInfo()
```

